---
title: "DataTable Basics Part 2"
author: "Gabriel Domingo"
date: "July 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Table Introduction

This document continues the introduction to Data.table from Part 1. Lets re-read the dataset.


```{r message=FALSE, include=FALSE}
require(tidyverse)
require(data.table)

dt1<-read_csv("flights14.csv")
head(dt1)
dt2<-fread("flights14.csv")
head(dt2)
```



## Task 1: Renaming Column Names

Another typical task is simply to rename columns. In tidyverse:

```{r}
dt1 %>% rename("departure delay"=dep_delay)
```
In data.table, we have `setnames`:

```{r}
setnames(dt2,'dep_delay','departure delay')
dt2 #permanently changing the name
```



## Task 2: Aggregations (Simple or by Group)

A typical task is to aggregate on a column. If that column is numeric, then aggregation would be getting a mean or other variable. If that column is discrete, we could get the counts per value.

Here are the codes for tidy:

```{r}
dt1 %>% group_by(origin) %>% summarise(count=n())
dt1 %>% group_by(origin) %>% summarise(count=n())
```

For continuous variables, we can calculate means per group. 

```{r}
dt1 %>% group_by(origin) %>% summarise(mean_dep_arr_delay=mean(arr_delay))

```

In data.table, this is also quite easy. This time we use the third part, the `by` part.

```{r}
dt2[,.(count=.N),by=.(origin)]
```

```{r}
dt2[,.(mean_dep_arr_delay=mean(arr_delay)),by=.(origin)]
dt2[,.(mean_dep_arr_delay=mean(arr_delay)),by="origin"] #equivalent

```

Notice the structure compared to $ \text{dt[i,j,by] }$. The `by` gives us the list of variables to calculate the aggregations. 


## Task 3: Creating New Variables as Functions of Another (1)

Another common task is to create a column, likely as a function of other variables. 

In tidy, this is the following:

```{r}
dt1 %>% group_by(origin) %>% mutate(mean_arr_delay=mean(arr_delay)) %>% select(mean_arr_delay)
```

For data.table, we use the `:=` operator:

```{r}
dt2[,mean_arr_delay:=mean(arr_delay),by=origin]
dt2[,c("origin","mean_arr_delay")] 
```

For both methods, what happened was a new variable was created, per value of origin. In this case, there are 3 unique values of origin, and these repeat throughout the dataset.
