---
title: "DataTable Basics Part 1"
author: "Gabriel Domingo"
date: "June 2020 (current version: April 2022)"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Table Introduction

This document explains how to do basics of data.table package.^[Current version 2: April 2022] Its an alternate way of doing the `tidyverse` way of data manipulation. In this tutorial, i will explain three tasks one usually does in tidyverse and give you the data.table equivalent.

But first, why do `data.table`? The best reason is performance. Please see this github page for performance benchmarks. The answer is that data.table is faster. As a user now for the past 2 years, i can attest to this.

https://github.com/Rdatatable/data.table/wiki/Benchmarks-:-Grouping


## Task 1: Reading Tables

Tidyverse has `readr` which allows one to import csvs. Data.table has that too; with fread.

```{r message=FALSE}
require(tidyverse)
require(data.table)

dt1<-read_csv("flights14.csv")
head(dt1)
```

```{r}
dt2<-fread("flights14.csv")
head(dt2)

```


Importing data this way ensures that the objects are a tibble for tidyverse, and data.table for Data.table. If you had a standard data.frame, you can convert to a tibble with `tibble()` and to data.table with `data.table()`. The type of data is also shown in Rmarkdown, but not in the pdf output. ^[Btw, i fully recommend using and learning Rmarkdown for reports generation.] Each importing command tries to import each column as the appropriate data type, although, the command enables the analyst to specify types if needed.

## Task 2: Filtering Rows

If we use tidy, we would use piping, and read from left to right. Use the `filter()` verb to enter a logical condition. You can enter multiple conditions the logical "and" `&`, or `|`, not `!`. For numerical, we can do `>`,`<` also. 

As an update, `R` has a native pipe in `|>` as a replacement/subst for the magrittr's `%>%`.

```{r}
dt1 %>% filter(origin=="JFK") 
dt1 %>% filter(origin!="JFK") 
dt1 %>% filter(origin!="JFK" & dep_delay>20)
```

```{r}
#using the native pipe, |>

dt1 |> filter(origin=="JFK") 

```

We can use data.table also. There is neat way to understand basic structure of a `data.table`. The verbs are directly used in this form:

$$ \text{dt[i,j,by] }$$
The `i` talks about rows, so generally operations on rows goes directly into this first position.

The `j` talks about columns, and `by` includes group-row operations.

Implementing the same notion for Data.table, we would put the filter directly:

```{r}
dt2[origin=="JFK",]
dt2[origin!="JFK" & dep_delay>20,]
```

### Task 3 : Selecting Columns

In Tidy, selecting columns/keeping columns uses the tidy verb `select`. Tidy usually requires verbs, and knowing the correct verbs. Meanwhile, data.table has fewer verbs, especially for these basic operations.

```{r}
dt1 %>% select(c("year","month","day"))
```
For data.table, we use the `j`, or the second component, making sure to add a "," at the start.

```{r}
dt2[,c("year","month","day")]
```

If you know base R, you would find this quite familiar. Here is another way to do it when you are doing it interactively. The `.()` format is a request for a list (this is a data.table thing, not important).

```{r}
dt2[,.(year,month,day)]
```


### Task 3 bonus -- selecting by Regular expression

We can select by in tidy using the `tidy select` verbs [Can you already see the problem with tidy] `contains` is a wrapper for other regex verbs in R.^[If you need more information on regex, we can do something separate. Regex can be quite useful.] 

```{r}
dt1 %>% select(contains("delay"),c("month","year","day"))
```

In data.table, we can also use functions to select columns.

```{r}
dt2[, .SD, .SDcols = patterns("delay|month|year|day",cols=names(dt2))]
dt2[, .SD, .SDcols = grep("delay|month|year|day",names(dt2),value=TRUE)]
```
This introduces the notion of `.SDcols` and `.SD`. SDCols allows the analyst to specify which columns she wants to work with. In fact, you can feed SDcols a character vector, or a regex command. 

To show the flexibility of data.table, i use two regex commands; the old staple `grep`, and the data.table convenience function `patterns`).

In the `j` part, you use the variable name `.SD` to tell data.table to use the results from `.SDcols`.

### Necessary Information on Datatypes and Creating a new Column

I might as well note here the datatypes one typically encounters. First is numeric (or integer) vector:

```{r}
str(dt1$dep_delay)
str(dt2$dep_delay)
# the $ symbol takes the data.table/tibble column and turns it into a vector
```

Next is Character vector. We can use regex (as above), or simply equality (also see above for filtering), to select characters.

```{r}
str(dt1$dest)
```

Next is factor. Think of a factor as ordered text. This is also important for regression later on, as a dummy variable.

```{r}
#get unique carriers:
unique(dt2$carrier)
#turn it into a factor

unique(dt2$carrier) |> factor()

```

In the above example, the carrier codes are ordered alphabetically. So, in a ggplot, they will also be ordered this way, and in a regression "AA" is the base case.

Finally, we have dates. This is actually fairly involved, but mostly we will be using daily dates for our work. There are several formats possible, but the most convenient one may be `date` command. We can manipulate dates as if they are numbers (find the diff), and when plotting, ggplot can order dates automatically.

We will have to construct dates as follows:

```{r}
as.Date("2020-01-01") #the format is year-month-day
```

Using the flights data, we can construct a new column which is a date column. As a bonus, we have demonstrated here the 2 ways in data.table and tidyverse to create a new column. We construct a date by pasting together the year, month and day information in the format date prefers.^[We can change the default format.]

```{r}
#using data.table to construct a new column:

dt2[,date:=as.Date(paste(year,month,day,sep="-"))]
dt2$date |> str()

# using the tidyverse way

dt1<-dt1 %>% mutate(date=as.Date(paste(year,month,day,sep="-")))
#in tidyverse, you need to overwrite the old dt1
dt1$date %>% str()
```

